---
title: 'Practical Machine Learning: Quiz 4'
author: "Fenton Taylor"
date: "November 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F)
```

##Problem 1 
Set the variable y to be a factor variable in both the training and test set. Then set the seed to 33833. Fit (1) a random forest predictor relating the factor variable y to the remaining variables and (2) a boosted predictor using the "gbm" method. Fit these both with the train() command in the caret package.

What are the accuracies for the two approaches on the test data set? What is the accuracy among the test set samples where the two methods agree? 

```{r problem1, cache=TRUE}
library(ElemStatLearn)
library(caret)
data(vowel.train)
data(vowel.test) 
vowel.test$y <- factor(vowel.test$y)
vowel.train$y <- factor(vowel.train$y)
set.seed(33833)

mod.rf <- train(y~.,method="rf", data=vowel.train)
mod.gbm <- train(y~., method="gbm", data=vowel.train, verbose=FALSE)

pred.rf <- predict(mod.rf, vowel.test)
pred.gbm <- predict(mod.gbm, vowel.test)

g <- ggplot(vowel.test, aes(pred.rf, pred.gbm, color=y))
g + geom_point(position = "jitter")

# RF Accuracy
acc.rf <- confusionMatrix(pred.rf, vowel.test$y)$overall[1]
# GBM Accuracy
acc.gbm <- confusionMatrix(pred.gbm, vowel.test$y)$overall[1]
# Agreement Accuracy
agree <- which((pred.rf==pred.gbm)==TRUE)
acc.agree <- confusionMatrix(pred.rf[agree], vowel.test$y[agree])$overall[1]
```

###Answer
* RF Accuracy = `r round(acc.rf, 4)`
* GBM Accuracy = `r round(acc.gbm, 4)`
* Agreement Accuracy = `r round(acc.agree, 4)`

## Problem 2
Load the Alzheimer's data using the following commands:

```{r problem2.load}
library(caret)
library(gbm)
set.seed(3433)
library(AppliedPredictiveModeling)
data(AlzheimerDisease)
adData = data.frame(diagnosis,predictors)
inTrain = createDataPartition(adData$diagnosis, p = 3/4)[[1]]
training = adData[ inTrain,]
testing = adData[-inTrain,]
```

Set the seed to 62433 and predict diagnosis with all the other variables using a random forest ("rf"), boosted trees ("gbm") and linear discriminant analysis ("lda") model. Stack the predictions together using random forests ("rf"). What is the resulting accuracy on the test set? Is it better or worse than each of the individual predictions? 

```{r problem2, cache=TRUE}
set.seed(62433)
mod.rf <- train(diagnosis~., method="rf", data = training)
mod.gbm <- train(diagnosis~., method="gbm", data = training, verbose=FALSE)
mod.lda <- train(diagnosis~., method="lda", data = training)

pred.rf <- predict(mod.rf, testing)
pred.gbm <- predict(mod.gbm, testing)
pred.lda <- predict(mod.lda, testing)
combo.df <- data.frame(pred.rf, pred.gbm, pred.lda, diagnosis = testing$diagnosis)
combo.mod <- train(diagnosis~., data=combo.df, method="rf")
combo.pred <- predict(combo.mod, combo.df)

accuracy.df <- data.frame(accuracy = c(confusionMatrix(pred.rf, testing$diagnosis)$overall[1],
                          confusionMatrix(pred.gbm, testing$diagnosis)$overall[1],
                          confusionMatrix(pred.lda, testing$diagnosis)$overall[1],
                          confusionMatrix(combo.pred, combo.df$diagnosis)$overall[1]))
row.names(accuracy.df) <- c("rf model", "gbm model", "lda model", "stacked model")
accuracy.df
```

###Answer
Stacked accuracy is `r round(accuracy.df[4,], 2)`, which is better than random forest (`r round(accuracy.df[1,], 2)`) and lda (`r round(accuracy.df[3,], 2)`), but about the same as boosting (`r round(accuracy.df[2,], 2)`).

##Problem 4

```{r problem4.load}
library(lubridate) # For year() function below
filename="C:/Users/sec/Desktop/Coursera/projects/Practical_Machine_Learning/Quizzes/gaData.csv"
if(!file.exists(filename)){
      download.file(url="https://d396qusza40orc.cloudfront.net/predmachlearn/gaData.csv",
                    destfile=filename)
}
dat = read.csv(filename)
training = dat[year(dat$date) < 2012,]
testing = dat[(year(dat$date)) > 2011,]
tstrain = ts(training$visitsTumblr)
```

Fit a model using the bats() function in the forecast package to the training time series. Then forecast this model for the remaining time points. For how many of the testing points is the true value within the 95% prediction interval bounds?

```{r problem4}
library(forecast)
mod <- bats(tstrain)
fcast <- forecast(mod, h=dim(testing)[1],level=95)
accuracy(fcast, testing$visitsTumblr)
acc <- sum(fcast$lower < testing$visitsTumblr & testing$visitsTumblr < fcast$upper)/dim(testing)[1]
acc
```

###Answer
Approximately `r round(acc,2)*100`% of the actual data fall within the 95% confidence interval. 

## Problem 3

Load the concrete data with the commands:

```{r problem3.load}
set.seed(3523)
library(AppliedPredictiveModeling)
data(concrete)
inTrain = createDataPartition(concrete$CompressiveStrength, p = 3/4)[[1]]
training = concrete[ inTrain,]
testing = concrete[-inTrain,]
```

Set the seed to 233 and fit a lasso model to predict Compressive Strength. Which variable is the last coefficient to be set to zero as the penalty increases? (Hint: it may be useful to look up ?plot.enet).

```{r problem3, cache=TRUE}
set.seed(233)
lasso.mod <- train(CompressiveStrength~., data=training, method="lasso")
plot.enet(lasso.mod$finalModel, xvar = "penalty", use.color = TRUE)
```

###Answer
Cement

##Problem 5

Set the seed to 325 and fit a support vector machine using the e1071 package to predict Compressive Strength using the default settings. Predict on the testing set. What is the RMSE? 

```{r prob5}
set.seed(325)
library(e1071)
mod.svm <- svm(CompressiveStrength~., data = training)
pred.svm <- predict(mod.svm, testing)

acc.table <- accuracy(pred.svm, testing$CompressiveStrength)
acc.table
```

###Answer
The RMSE is `r acc.table[,"RMSE"]`.

The End